The Concurrent Calculi Formalisation Benchmark

### 2024-02-23 challenges/preliminaries.tex
Remove old comments.

The syntax of processes includes:
the process \( \Pend \) or \emph{inaction}, a process which can do nothing. The process \( \Ppar{P}{Q} \) is the \emph{parallel composition} of process \( P \) and process \( Q \).
The two components can proceed independently of each other, or they can interact via shared names.
%The process \( \Pchoice{P}{Q} \) is a non-deterministic \emph{choice} between continuing as the process \( P \) or as the process \( Q \).

For communication, processes include \emph{input} and \emph{output}, whose
signature signature depends on the calculus being value-passing or name-passing.
We use here the metavariables $c,k$ to abstract over this choice
--- i.e.~$c$ may be either a value or a variable or a name, whereas $k$ may
be a variable or a name.
The process \( \Pout{x}{c}{P} \) is an \emph{output}, which can send
\( c \) via \( x \), then continue as \( P \).  % The intention is that
% the value \( v \) must be a base value when it is actually sent, and
% this will be enforced in the semantics later on
The process \( \Pin{x}{k}{P} \) is an \emph{input}, which can receive a $c$
via \( x \), then continue as \( P \) with the received element
substituted for \( k \).  The input operator thus
binds \( k \) in \( P \).

@@

We use the notation \( \freenames{P} \) to denote the set of
names that occur free,
% (i.e.\ not bound by a restriction) in \( P \),
\( \boundnames{P} \) to denote the set of names that occur bound
% (by a restriction)
in \( P \) and  \( \freevars{P} \)
to denote the set of variables that occur free
% (i.e.\ not bound by an input)
in \( P \).  We use the notation \( \boundvars{P} \)
for the set of variables that occur bound
% (by an input)
in \( P \).  We use the notation \( \subst{P}{a}{l} \) to denote
the process \( P \) with base value \( a \) substituted for variable
\( l \). Similarly, \( \subst{P}{x}{y} \) denotes the process
\( P \) with name \( x \) substituted for name \( y \).
We use the notation \( \applysubst{\sigma}{P} \) to denote the process \( P \) with a finite number of arbitrary substitutions applied to it.

@@

A \emph{context} is obtained by taking a process and replacing a single occurrence of \( \Pend \) in it with the special \emph{hole} symbol \( \ctxhole \).
As a convention, we do \emph{not} identify \( \alpha \)-convertible contexts.
%
A context acts as a function between processes:
a context \( C \) can be \emph{applied} to a process \( P \), written \( \applyctx{C}{P} \), by replacing the hole in C by \( P \), thus obtaining another process.
The replacement is literal, so names and variables that are free in \( P \) can become bound in \( \applyctx{C}{P} \).

>>>>

The syntax of processes includes:
the process \( \Pend \) or \emph{inaction}, a process which can do nothing. The process \( \Ppar{P}{Q} \) is the \emph{parallel composition} of process \( P \) and process \( Q \).
The two components can proceed independently of each other, or they can interact via shared names.

For communication, processes include \emph{input} and \emph{output}, whose
signature signature depends on the calculus being value-passing or name-passing.
We use here the metavariables $c,k$ to abstract over this choice
--- i.e.~$c$ may be either a value or a variable or a name, whereas $k$ may
be a variable or a name.
The process \( \Pout{x}{c}{P} \) is an \emph{output}, which can send
\( c \) via \( x \), then continue as \( P \).
The process \( \Pin{x}{k}{P} \) is an \emph{input}, which can receive a $c$
via \( x \), then continue as \( P \) with the received element
substituted for \( k \).  The input operator thus
binds \( k \) in \( P \).

@@

We use the notation \( \freenames{P} \) to denote the set of
names that occur free,
\( \boundnames{P} \) to denote the set of names that occur bound
in \( P \) and  \( \freevars{P} \)
to denote the set of variables that occur free
in \( P \).  We use the notation \( \boundvars{P} \)
for the set of variables that occur bound
in \( P \).  We use the notation \( \subst{P}{a}{l} \) to denote
the process \( P \) with base value \( a \) substituted for variable
\( l \). Similarly, \( \subst{P}{x}{y} \) denotes the process
\( P \) with name \( x \) substituted for name \( y \).
We use the notation \( \applysubst{\sigma}{P} \) to denote the process \( P \) with a finite number of arbitrary substitutions applied to it.

@@

A \emph{context} is obtained by taking a process and replacing a single occurrence of \( \Pend \) in it with the special \emph{hole} symbol \( \ctxhole \).
As a convention, we do \emph{not} identify \( \alpha \)-convertible contexts.
A context acts as a function between processes:
a context \( C \) can be \emph{applied} to a process \( P \), written \( \applyctx{C}{P} \), by replacing the hole in C by \( P \), thus obtaining another process.
The replacement is literal, so names and variables that are free in \( P \) can become bound in \( \applyctx{C}{P} \).

### 2024-02-23 short-paper/short-paper.tex
Remove LaTeX that modifies the page layout.
Remove old comments.

\documentclass[runningheads]{llncs}
\synctex=1
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{mathpartir}
\usepackage{cite}
\usepackage{bm}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{hyperref}
\usepackage{microtype}
\usepackage[inline]{enumitem}
\usepackage{color}
\renewcommand\UrlFont{\color{blue}\rmfamily}

\usepackage[status=draft]{fixme}
\fxsetup{theme=color}
% \setlength{\textwidth}{12.4cm}
\setlength{\textheight}{19.9cm}

@@

\begin{abstract}
  POPLMark and POPLMark Reloaded sparked a flurry of work on machine-checked proofs, and fostered the adoption of proof mechanisation in programming language research.
  Both challenges were purposely limited in scope, and their benchmark challenges do not address concurrency-related issues.
  %
  We propose a new collection of benchmark challenges focused on the difficulties that typically arise when mechanising formal models of concurrent and distributed programming languages, such as process calculi.
  Our benchmark challenges address three key topics: linearity, scope extrusion, and coinductive reasoning.
  The goal of this new challenge is to clarify, compare, and advance the state of the art, fostering the adoption of proof mechanisation in future research on concurrency.

@@

We have begun collecting solutions to our challenges on our website:
%
\begin{center}
  \url{https://concurrentbenchmark.github.io/}
\end{center}
%
In the future, we intend to use the website to promote best practices and tutorials derived from solutions to our challenges.
We encourage readers to try the challenges using their favourite techniques, and to send us their solutions.

@@

In mechanised meta-theory, addressing linearity means choosing an appropriate
representation of the linear typing context.  While a linear context is perhaps
best seen as a multiset, most proof assistants have better support for list
processing.  The latter representation is intuitive, but may require
establishing a large number of technical lemmata that are orthogonal to the
problem under study (in our case, proving type safety for
session types).  In this space, several designs are possible: one can label
occurrences of resources to constrain their usage (\eg~\cite{CicconeP20}), or impose a multiset structure over lists
(\eg~\cite{Danielsson12,ChaudhuriLR19}). Alternatively, contexts can be
implemented as finite maps (as in \cite{Castro2020}), whose operations are
sensitive to a linear discipline. In all these cases, the effort required to
develop the infrastructure is significant.
%\todo{Consider adding material on linearity predicates}
%
One alternative strategy is to bypass the problem of context splitting
by adopting familiar ideas from algorithmic linear type checking. This
approach, referred to as ``typing with leftovers,'' is exemplified
in~\cite{DBLP:conf/forte/ZalakainD21}.    Similarly, one
could eliminate context splitting by delegating linearity checks to a
\emph{linear predicate} defined on the process structure (\eg~\cite{BP23}). These checks 
serve as additional conditions within the typing rule. Whatever the choice,
list-based encodings can be refined to be intrinsically-typed, if the
proof assistant supports dependent types
(see~\cite{Thiemann2019,CicconeP20,RouvoetPKV20}).
%

@@

Given those peculiarities, the syntax and semantics of $\pi$-calculi have been 
mechanised from an early age (see~\cite{Melham1994}) with many proof
assistants and in so many encoding styles that it would take too long to
review.
Despite this, almost all of these mechanisations still rely on ad-hoc solutions to accommodate scope extrusion.
They range from concrete encodings based on named syntax
\cite{Melham1994} to basic de Bruijn~\cite{Hirschkoff1997, Perera2018} and locally-nameless
representation %~\cite{Chargueraud2012}
(\eg~\cite{Castro2020}).
Nominal approaches are also common (see~\cite{Bengtson2009}), but they
may be problematic in proof assistants based on dependent types.
The case study \cite{AmbalLS21} compares four approaches to encoding
binders in Coq as a first step towards developing tools for working
with higher-order process calculi.  They found that working directly
with de~Bruijn indices was easiest since the existing
approaches developed for $\lambda$-calculus binders worked poorly 
with scope extrusion.

@@

When reasoning over bisimilarity many authors rely on the native
 coinduction offered by the chosen proof assistants
\cite{Bengtson2016,Kahsai2008,Thiemann2019,Gay2020}, while others prefer a
more ``set-theoretic'' approach
\cite{Hirschkoff1997,Bengtson2009,Maksimovic2015,Pohjola2022}. Some
use both and establish an internal adequacy~\cite{Honsell2001}. Few
extend the proof assistant foundations to allow, \eg, reasoning about
bisimilarity up-to~\cite{ChaudhuriCM15}.
% While many proof assistants support
% coinductive techniques,   but they support subtly different formalisms that enable different techniques, and some systems provide more than one abstraction to use coinduction.
% We thus see a variety of different formalisation choices for
% bisimulations, sometimes ad-hoc for the specific result.

% \todo{Reorganize the rest of this section to be more coherent}
% Isabelle, Coq, and Abella seamlessly support coinductive definitions and reasoning.
% Agda and Beluga support it through the convenient abstraction of co-patterns~\cite{Abel2013}.
% Newer entrants to the field like the Lean proof assistant
% are working on adding a notion of coinduction to their system~\cite{Avigad2019}.\alb{I kinda object to ``seamlessly''. In Coq it's a mess, in Abella very very limited, sized types in Agda are shaky. I guess Isabelle is better}
% =======
% Process calculi often include constructs that allow processes to have infinite behaviours.
% Coinduction is a fundamental technique for defining and reasoning about infinite objects, and it can thus be used to reason about these behaviours.
% While many proof assistants support coinductive techniques, using them is notoriously difficult, and representations and proof techniques vary widely.
% The crux of our challenge is the effective use of coinductive up-to techniques.
% The intention is that it should be easy to use the bisimulation relation and its properties, but this may be difficult with some mechanisation approaches.

% The goal of our challenge on coinductive reasoning is proving that strong barbed bisimilarity can be turned into a congruence by making it sensitive to substitution and parallel composition.
% The setting for our challenge is an untyped \(\pi\)-calculus with replication of processes enabling infinite behaviours.
% We do not include delegation in the calculus since it is orthogonal to our primary aim of exploring coinductive proof techniques.
% We base our definition of bisimilarity on a labelled transition system semantics and an observability predicate describing the communication steps available to a process.
% The introduction of strong barbed bisimulation is one of the first steps when studying the observable behaviour of process calculi, both in classical textbooks (\eg \cite{picalcbook}) and in existing mechanisations.
% Though weak barbed congruence is a more common behavioural equivalence, we prefer strong equivalences to simplify the theory by avoiding the need to abstract over the number of internal transitions in a trace.

% Coinduction is widely supported in proof assistants, but they support subtly different formalisms that enable different techniques, and some systems provide more than one abstraction to use coinduction.
% We thus see a variety of different formalisation choices for bisimulations, sometimes ad-hoc for the specific result.

% \todo{Reorganize the rest of this section to be more coherent}
% Isabelle, Coq, and Abella seamlessly support coinductive definitions and reasoning.
% Agda and Beluga support it through the convenient abstraction of co-patterns~\cite{Abel2013}.
% Newer entrants to the field like the Lean proof assistant
% are working on adding a notion of coinduction to their system~\cite{Avigad2019}.\alb{I kinda object to ``seamlessly''. In Coq it's a mess, in Abella very very limited, sized types in Agda are shaky. I guess Isabelle is better}

@@

% - Going beyond the challenge problems
%   - We do not aim to be comprehensive, but the techniques should also be useful for other things
%   - Our challenges are independent, but they can also be combined
%   - Our challenges are basic textbook theory, so they can form the basis for many extensions
%   - Our aim is to see a future where ``the basics'' are as trivial in mechanisations as they are on paper

% We previously promise to discuss combinations of techniques and applications to more elaborate schemes
%\section{Going Beyond the Challenge Problems}\label{sec:going-beyond}
\section{Future Work and Conclusions}\label{sec:going-beyond}
Our benchmark challenges do not cover all issues in the field, but focus on the fundamental aspects of linearity, scope extrusion, and coinduction.
Many mechanisations need to combine techniques to handle several of these aspects at once, and some may also need to handle aspects that are not covered by our benchmark challenges at all.

%%% Combinations
Combining techniques for mechanising the fundamental aspects covered in our benchmark is often not trivial.
While we have chosen to focus on the aspects individually to simplify the challenges, we are of course also interested in exploring how techniques interact.
Some authors have already described the difficulties that combining techniques may induce in a mechanisation, \eg~\cite{DBLP:conf/forte/ZalakainD21}.\todo{Add more references here?}

%%% Other aspects
Much current research on concurrent calculi includes aspects that are not covered by our benchmark challenges.
Features that are in wide use, but not covered by our benchmark, include choice constructs and recursion.
Some interesting research topics that build on the fundamental aspects in our benchmark include multiparty session types~\cite{10.1145/2827695}, choreographies~\cite{Carbone2013}, higher-order calculi~\cite{Hirsch2022}, conversation types~\cite{DBLP:journals/tcs/CairesV10}, psi-calculi~\cite{lmcs:696}, and encodings between different calculi~\cite{DBLP:journals/iandc/Gorla10,DBLP:journals/iandc/DardhaGS17}.
Their meta-theory includes aspects---\eg well-formedness conditions on global types, partiality of end-point projection functions, \etc---that we do not address.

@@

%%% Conclusion
Ultimately, the fundamental aspects covered by our benchmark serve as the building blocks for most current research on concurrenct calculi.
It is our hope and aim that exploring solutions to our challenges will move the community closer to a future where the key basic proof techniques for concurrent calculi are as easy to mechanise as they are to write on paper.

>>>>

\documentclass[runningheads]{llncs}
\synctex=1
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{mathpartir}
\usepackage{cite}
\usepackage{bm}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{microtype}
\usepackage[inline]{enumitem}
\usepackage{color}

@@

\begin{abstract}
  POPLMark and POPLMark Reloaded sparked a flurry of work on machine-checked proofs, and fostered the adoption of proof mechanisation in programming language research.
  Both challenges were purposely limited in scope, and their benchmark challenges do not address concurrency-related issues.
  We propose a new collection of benchmark challenges focused on the difficulties that typically arise when mechanising formal models of concurrent and distributed programming languages, such as process calculi.
  Our benchmark challenges address three key topics: linearity, scope extrusion, and coinductive reasoning.
  The goal of this new challenge is to clarify, compare, and advance the state of the art, fostering the adoption of proof mechanisation in future research on concurrency.

@@

We have begun collecting solutions to our challenges on our website:
\begin{center}
  \url{https://concurrentbenchmark.github.io/}
\end{center}
In the future, we intend to use the website to promote best practices and tutorials derived from solutions to our challenges.
We encourage readers to try the challenges using their favourite techniques, and to send us their solutions.

@@

In mechanised meta-theory, addressing linearity means choosing an appropriate
representation of the linear typing context.  While a linear context is perhaps
best seen as a multiset, most proof assistants have better support for list
processing.  The latter representation is intuitive, but may require
establishing a large number of technical lemmata that are orthogonal to the
problem under study (in our case, proving type safety for
session types).  In this space, several designs are possible: one can label
occurrences of resources to constrain their usage (\eg~\cite{CicconeP20}), or impose a multiset structure over lists
(\eg~\cite{Danielsson12,ChaudhuriLR19}). Alternatively, contexts can be
implemented as finite maps (as in \cite{Castro2020}), whose operations are
sensitive to a linear discipline. In all these cases, the effort required to
develop the infrastructure is significant.
One alternative strategy is to bypass the problem of context splitting
by adopting familiar ideas from algorithmic linear type checking. This
approach, referred to as ``typing with leftovers,'' is exemplified
in~\cite{DBLP:conf/forte/ZalakainD21}.    Similarly, one
could eliminate context splitting by delegating linearity checks to a
\emph{linear predicate} defined on the process structure (\eg~\cite{BP23}). These checks 
serve as additional conditions within the typing rule. Whatever the choice,
list-based encodings can be refined to be intrinsically-typed, if the
proof assistant supports dependent types
(see~\cite{Thiemann2019,CicconeP20,RouvoetPKV20}).

@@

Given those peculiarities, the syntax and semantics of $\pi$-calculi have been 
mechanised from an early age (see~\cite{Melham1994}) with many proof
assistants and in so many encoding styles that it would take too long to
review.
Despite this, almost all of these mechanisations still rely on ad-hoc solutions to accommodate scope extrusion.
They range from concrete encodings based on named syntax
\cite{Melham1994} to basic de Bruijn~\cite{Hirschkoff1997, Perera2018} and locally-nameless
representation (\eg~\cite{Castro2020}).
Nominal approaches are also common (see~\cite{Bengtson2009}), but they
may be problematic in proof assistants based on dependent types.
The case study \cite{AmbalLS21} compares four approaches to encoding
binders in Coq as a first step towards developing tools for working
with higher-order process calculi.  They found that working directly
with de~Bruijn indices was easiest since the existing
approaches developed for $\lambda$-calculus binders worked poorly 
with scope extrusion.

@@

When reasoning over bisimilarity many authors rely on the native
 coinduction offered by the chosen proof assistants
\cite{Bengtson2016,Kahsai2008,Thiemann2019,Gay2020}, while others prefer a
more ``set-theoretic'' approach
\cite{Hirschkoff1997,Bengtson2009,Maksimovic2015,Pohjola2022}. Some
use both and establish an internal adequacy~\cite{Honsell2001}. Few
extend the proof assistant foundations to allow, \eg, reasoning about
bisimilarity up-to~\cite{ChaudhuriCM15}.

@@

\section{Future Work and Conclusions}\label{sec:going-beyond}
Our benchmark challenges do not cover all issues in the field, but focus on the fundamental aspects of linearity, scope extrusion, and coinduction.
Many mechanisations need to combine techniques to handle several of these aspects at once, and some may also need to handle aspects that are not covered by our benchmark challenges at all.

Combining techniques for mechanising the fundamental aspects covered in our benchmark is often not trivial.
While we have chosen to focus on the aspects individually to simplify the challenges, we are of course also interested in exploring how techniques interact.
Some authors have already described the difficulties that combining techniques may induce in a mechanisation, \eg~\cite{DBLP:conf/forte/ZalakainD21}.\todo{Add more references here?}

Much current research on concurrent calculi includes aspects that are not covered by our benchmark challenges.
Features that are in wide use, but not covered by our benchmark, include choice constructs and recursion.
Some interesting research topics that build on the fundamental aspects in our benchmark include multiparty session types~\cite{10.1145/2827695}, choreographies~\cite{Carbone2013}, higher-order calculi~\cite{Hirsch2022}, conversation types~\cite{DBLP:journals/tcs/CairesV10}, psi-calculi~\cite{lmcs:696}, and encodings between different calculi~\cite{DBLP:journals/iandc/Gorla10,DBLP:journals/iandc/DardhaGS17}.
Their meta-theory includes aspects---\eg well-formedness conditions on global types, partiality of end-point projection functions, \etc---that we do not address.

@@

Ultimately, the fundamental aspects covered by our benchmark serve as the building blocks for most current research on concurrenct calculi.
It is our hope and aim that exploring solutions to our challenges will move the community closer to a future where the key basic proof techniques for concurrent calculi are as easy to mechanise as they are to write on paper.
